Type theories as quotient inductive-recursive types

### 2025-02-14 tex/main.tex
Add points discussed during the meeting

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

\subsection{Inductive-recursive types}

@@

@@

\subsection{The standard model} \label{subsec:std-model}
The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{align*}
  \Ctx^M\;\nameless          & \defeq \Set              & \Sub^M\;\Gamma^M\;\Delta^M\;\nameless & \defeq \Gamma^M \to \Delta^M \\
  \Ty^M\;\Gamma^M\;\nameless & \defeq \Gamma^M \to \Set & \Tm^M\;\Gamma^M\;A^M\;\nameless & \defeq (\gamma : \Gamma^M) \to A^M\;\gamma
\end{align*}
%with the additional methods to its QII counterpart:
%\begin{align*}
%  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
%  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
%\end{align*}
%where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
%
We once again refer to our Agda formalisation for details.
We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.

@@

\LT{Mention the NbE for SC here}

@@

\Fredrik{TODO}

@@

\section{Formal definitions in Agda}
\end{document}

>>>>

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, \Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

\subsection{Inductive-recursive types}
\LT[noinline]{Shorten this section}
\LT[noinline]{the recursion part of IR is not given by the elimination rule}

@@

\LT[noinline]{remark that recursions in IR are not defined by elimination rules}

@@

\subsection{The standard model} \label{subsec:std-model}
The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{align*}
  \Ctx^M\;\nameless          & \defeq \Set              & \Sub^M\;\Gamma^M\;\Delta^M\;\nameless & \defeq \Gamma^M \to \Delta^M \\
  \Ty^M\;\Gamma^M\;\nameless & \defeq \Gamma^M \to \Set & \Tm^M\;\Gamma^M\;A^M\;\nameless & \defeq (\gamma : \Gamma^M) \to A^M\;\gamma
\end{align*}
%with the additional methods to its QII counterpart:
%\begin{align*}
%  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
%  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
%\end{align*}
%where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
%
We once again refer to our \Agda formalisation for details.
We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.

@@

\LT[noinline]{Mention the NbE for SC here?}

@@

\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set &
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}

\begin{alignat*}{10}
  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  [\blank]_{\Ty}^M\blank & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
  \idS^M          & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  \blank;^M\blank & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma\;\Delta^M\;\sigma) \to (\tau^M : \Sub^M\;\Delta^M\;\Theta^M\;\tau) \to \Sub^M\;\Gamma^M\;\Delta^M\;(\sigma ; \tau) \\
  \pi_1^M         & : \implicit{\dots}\; && \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma \to \Sub^M\;\Gamma^M\;\Delta^M\;(\pi_1\sigma) \\
  \pi_2^M         & : \implicit{\dots}\; && (\sigma^M : \Sub^M\;\Gamma^M\;(\Delta^M ,_{\Ctx}^M A^M)\;\sigma) \to \Tm^M\;\Gamma^M\;([ \pi_1^M\,\sigma^M ]^M_{\Ty}\,A^M)\;(\pi_2\,\sigma) \\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)\;([\sigma]\;t) \\
  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
  \UU^M & : \implicit{\dots}\; && \Ty^M\;\Gamma^M\;\UU \\
  \El^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;\UU^M\;u \to \Ty^M\;\Gamma^M\;(\El\,u) \\
  \Pi^M & : \implicit{\dots}\; && (A^M : \Ty^M\;\Gamma^M\;A) \to \Ty^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B \to \Ty^M\;\Gamma^M\;(\Pi\,A\,B)  \\
  \lambda^M & : \implicit{\dots}\; && \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;t \to \Tm^M\;\Gamma^M\;(\Pi^M\;A^M\;B^M)\;(\lambda\,t) \\
  \mathsf{app}^M & : \implicit{\dots}\; && \Tm^M\;\Gamma^M\;(\Pi^M\,A^M\,B^M)\;t \to \Tm^M\;(\Gamma^M ,_{\Ctx}^M A^M)\;B^M\;(\mathsf{app}\,t) \\
\end{alignat*}
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M
\end{alignat*}
\LT{To be finished}

@@

\end{document}
